---
title: "Chapter 2 Probability"
format: html
editor: visual
---

```{r}
#| include: false
library(tidyverse)
```

# 2.1 Probability basics

-   An *experiment* is a process that produces an *observation*

-   An *outcome* is a possible observation

-   The set of all possible outcomes is called the *sample space*

-   An *event* is a subset of the sample space

-   A *trial* is a single running of an experiment

Let $A$ and $B$ be *events in the sample space* $S$.

-   $P(\emptyset)=0$

-   If $A$ and $B$ are *disjoint*, then $P(A \cup B)=P(A) + P(B)$

-   If $A \subset B$, then $P(A) \leq P(B)$.

-   $0 \leq P(A) \leq 1$.

-   $P(A) = 1 - P (\overline{A})$.

-   $P(A-B)=P(A)-P(A\cap B)$.

-   $P(A \cup B)=P(A)+P(B)-P(A \cap B)$.

<br>

The *odds* of an event $E$ are $\frac{P(E)}{1-P(E)}$.

# 2.2 Simulations

## 2.2.1 Simulation with `sample`

Get a random number from 1 to 10:

```{r}
sample(x = 1:10, size = 1)
```

Define how many random numbers we want:

```{r}
sample(x = 1:10, size = 8)
```

If we sample for more than 10 different numbers, we get an error:

```{r}
#| error: true
sample(x = 1:10, size = 30)
```

To sample with replacement:

```{r}
sample(x = 1:10, size = 20, replace = TRUE)
```

Example: US blood types

```{r}
bloodtypes <- c("O", "A", "B", "AB")
bloodprobs <- c(0.45, 0.40, 0.11, 0.04)
sample(x = bloodtypes, size = 30, prob = bloodprobs, replace = TRUE)
```

A large sample reproduces the original probabilities with reasonable accuracy:

```{r}
sim_data <- sample(
  x = bloodtypes, size = 10000,
  prob = bloodprobs, replace = TRUE
)
table(sim_data)
table(sim_data) / 10000
rstatix::freq_table(sim_data)
```

## 2.2.2 Using simulation to estimate probabilities

Three-step process:

1.  Simulate the experiment many times to produce a vector of outcomes
2.  Test if the outcomes are in the event to produce a vector of TRUE/FALSE
3.  Compute the `mean` of the TRUE/FALSE vector to compute the probability estimate

```{r}
# Simulate 10,000 rolls of 2 dice
die1 <- sample(x = 1:6, size = 10000, replace = TRUE)
die2 <- sample(x = 1:6, size = 10000, replace = TRUE)

# Add the two dice
sumDice <- die1 + die2
```

```{r}
head(die1)
head(die2)
head(sumDice)
```

Let $E$ be the event "the sum of the dice is 6", and $F$ be the event "at least one of the dice is a 2".

```{r}
eventE <- sumDice == 6
head(eventE)
```

```{r}
eventF <- die1 == 2 | die2 == 2
head(eventF)
```

From theory, $P(E)=\frac{5}{36}\approx0.139$, and $P(F)=\frac{11}{36}\approx0.306$.

```{r}
mean(eventE)  # P(E)
mean(eventF)  # P(F)
```

To estimate $P(E \cap F) = \frac{2}{36} \approx 0.056$:

```{r}
mean(eventE & eventF)
```

To estimate $P(E \cup F) \approx 0.389$:

```{r}
mean((sumDice == 6) | (die1 == 2 | die2 == 2))
```

**Bernouilli's theorem:**

Let $E$ be an event with probability $p$. Let $m_n$ be the number of times that $E$ occurs in $n$ repeated trials, where we assume the outcome of trials do not affect the outcome of other trials. Then

$$
\lim_{x\to\infty} \frac{m_n}{n}=p
$$

## 2.2.3 Using `replicate` to repeat experiments

Estimate the probability that the sum of seven dice is larger than 30.

Simulate the event once:

```{r}
dice <- sample(x = 1:6, size = 7, replace = TRUE)
sum(dice) > 30
```

Using `replicate` repeats the experiment many times:

```{r}
replicate(n = 20,
          expr = {
            dice <- sample(x = 1:6, size = 7, replace = TRUE)
            sum(dice) > 30
          })
```

To compute the probability of the event, we replicate 10,000 times for a reasonably accurate estimate:

```{r}
event <- replicate(
  n = 10000,
  expr = {
    dice <- sample(x = 1:6, size = 7, replace = TRUE)
    sum(dice) > 30
    })

mean(event)
```

Recommendations:

1.  Write code that performs the experiment a single time
2.  Replicate the experiment a small number of times and check the results
3.  Replicate the experiment a large number of times and store the result
4.  Compute probability using `mean(event)`

Example: three dice are thrown. Estimate the probability that the largest value is a four.

```{r}
# One trial
die_roll <- sample(x = 1:6, size = 3, replace = TRUE)
max(die_roll) == 4
```

```{r}
# A few trials
replicate(n = 20, 
          expr = {
            die_roll <- sample(x = 1:6, size = 3, replace = TRUE)
            max(die_roll) == 4
          })
```

```{r}
# Perform many trials and compute the probability of the event
event <- replicate(
  n = 10000, 
  expr = {
    die_roll <- sample(x = 1:6, size = 3, replace = TRUE)
    max(die_roll) == 4
    })
mean(event)
```

Example: two dice are rolled. What is the probability that the sum is equal to 6?

In theory: $P(E)=\frac{5}{36} \approx 0.138888$.

```{r}
# 10 trials
mean(
  replicate(
    n = 10,
    expr = {
      sum(sample(x = 1:6, size = 2, replace = TRUE)) == 6
    }))
```

```{r}
# 200 trials
mean(
  replicate(
    n = 200,
    expr = {
      sum(sample(x = 1:6, size = 2, replace = TRUE)) == 6
    }))
```

```{r}
# 10,000 trials
mean(
  replicate(
    n = 10000,
    expr = {
      sum(sample(x = 1:6, size = 2, replace = TRUE)) == 6
    }))
```

Estimate the probability of $E$ for simulations ranging from 100 trials to 10,000 trials.

```{r}
num_trials <- seq(from = 100, to = 10000, by = 100)

res <- vector(mode = "numeric", length = length(num_trials))

for (i in 1:length(num_trials)) {
  res[i] <- mean(
    replicate(
      n = num_trials[i],
      expr = {
        sum(sample(x = 1:6, size = 2, replace = TRUE)) == 6
    }))
}
```

Illustration of the law of large numbers: probability estimates from simulation converge to the true probability as the number of trials increases.

```{r}
tibble(
  trials = num_trials,
  estimated_probability = res
) |> 
  ggplot() +
  geom_line(aes(x = num_trials,
                y = estimated_probability)) +
  geom_hline(yintercept = 5/36)
```

Example: a fair coin is repeatedly tossed. Estimate the probability that you observe heads for the third time on the 10th toss.

```{r}
# Simulate one outcome
coinToss <- sample(x = c("H", "T"), size = 10, replace = TRUE)
coinToss
```

```{r}
# Count the number of heads and check whether it is equal to three
sum(coinToss == "H")
sum(coinToss == "H") == 3
```

```{r}
# Make sure we had only two heads in the first nine tosses
coinToss[1:9]
sum(coinToss[1:9] == "H") == 2
```

```{r}
# Both of those have to be true in order for the event to occur
sum(coinToss == "H") == 3 & sum(coinToss[1:9] == "H") == 2
```

Put this inside `replicate` and compute the probability

```{r}
event <- replicate(
  n = 10000,
  expr = {
    coinToss <- sample(x = c("H", "T"), size = 10, replace = TRUE)
    (sum(coinToss == "H") == 3 & sum(coinToss[1:9] == "H") == 2)
  })

mean(event)
```

Example: the birthday problem.

Estimate the probability that out of 25 randomly selected people, at least two will have the same birthday. Assume that all birthdays are equaly likely, except that none are leap-day babies.

An outcome for this experiment is 25 randomly selected birthdays, and the event $B$ is that at least two birthdays are the same.

```{r}
# Simulate the experiment
set.seed(42)
birthdays <- sample(x = 1:365, size = 25, replace = TRUE)
birthdays
```

```{r}
# Check for duplicates
anyDuplicated(birthdays)
which(birthdays == birthdays[8])
```

```{r}
eventB <- replicate(
  n = 10000,
  expr = {
    birthdays <- sample(x = 1:365, size = 25, replace = TRUE)
    anyDuplicated(birthdays) > 0
    })
mean(eventB)
```

Example: three numbers are picked uniformly at random from the interval $(0, 1)$. What is the probability that a triangle can be formed whose side-lengths are the three numbers that you chose?

We need to check whether the sum of the two smaller numbers is larger than the largest number.

```{r}
event <- replicate(
  n = 10000,
  expr = {
    x <- sort(runif(n = 3, min = 0, max = 1))
    sum(x[1:2]) > x[3]
  })
mean(event)
```

Example: if you buy a bag that contains 35 M&M's, what is the probability that it will contain exactly 9 Blue and 5 Red M&M's?

```{r}
mm.colors <- c("Blue", "Orange", "Green", "Yellow", "Red", "Brown")
mm.probs <- c(25, 25, 12.5, 12.5, 12.5, 12.5)
```

```{r}
bag <- sample(
  x = mm.colors,
  size = 35,
  replace = TRUE,
  prob = mm.probs
)
sum(bag == "Blue")
```

```{r}
event <- replicate(
  n = 10000,
  expr = {
    bag <- sample(
      x = mm.colors,
      size = 35,
      replace = TRUE,
      prob = mm.probs
      )
    sum(bag == "Blue") == 9 & sum(bag == "Red") == 5
  })
mean(event)
```

# 2.3 Conditional probability and independence
