---
title: "Chapter 2 Probability"
format: html
editor: visual
---

# 2.1 Probability basics

-   An *experiment* is a process that produces an *observation*

-   An *outcome* is a possible observation

-   The set of all possible outcomes is called the *sample space*

-   An *event* is a subset of the sample space

-   A *trial* is a single running of an experiment

Let $A$ and $B$ be *events in the sample space* $S$.

-   $P(\emptyset)=0$

-   If $A$ and $B$ are *disjoint*, then $P(A \cup B)=P(A) + P(B)$

-   If $A \subset B$, then $P(A) \leq P(B)$.

-   $0 \leq P(A) \leq 1$.

-   $P(A) = 1 - P (\overline{A})$.

-   $P(A-B)=P(A)-P(A\cap B)$.

-   $P(A \cup B)=P(A)+P(B)-P(A \cap B)$.

<br>

The *odds* of an event $E$ are $\frac{P(E)}{1-P(E)}$.

# 2.2 Simulations

## 2.2.1 Simulation with `sample`

Get a random number from 1 to 10:

```{r}
sample(x = 1:10, size = 1)
```

Define how many random numbers we want:

```{r}
sample(x = 1:10, size = 8)
```

If we sample for more than 10 different numbers, we get an error:

```{r}
#| error: true
sample(x = 1:10, size = 30)
```

To sample with replacement:

```{r}
sample(x = 1:10, size = 20, replace = TRUE)
```

Example: US blood types

```{r}
bloodtypes <- c("O", "A", "B", "AB")
bloodprobs <- c(0.45, 0.40, 0.11, 0.04)
sample(x = bloodtypes, size = 30, prob = bloodprobs, replace = TRUE)
```

A large sample reproduces the original probabilities with reasonable accuracy:

```{r}
sim_data <- sample(
  x = bloodtypes, size = 10000,
  prob = bloodprobs, replace = TRUE
)
table(sim_data)
table(sim_data) / 10000
rstatix::freq_table(sim_data)
```

## 2.2.2 Using simulation to estimate probabilities

Three-step process:

1.  Simulate the experiment many times to produce a vector of outcomes
2.  Test if the outcomes are in the event to produce a vector of TRUE/FALSE
3.  Compute the `mean` of the TRUE/FALSE vector to compute the probability estimate

```{r}
# Simulate 10,000 rolls of 2 dice
die1 <- sample(x = 1:6, size = 10000, replace = TRUE)
die2 <- sample(x = 1:6, size = 10000, replace = TRUE)

# Add the two dice
sumDice <- die1 + die2
```

```{r}
head(die1)
head(die2)
head(sumDice)
```

Let $E$ be the event "the sum of the dice is 6", and $F$ be the event "at least one of the dice is a 2".

```{r}
eventE <- sumDice == 6
head(eventE)
```

```{r}
eventF <- die1 == 2 | die2 == 2
head(eventF)
```

From theory, $P(E)=\frac{5}{36}\approx0.139$, and $P(F)=\frac{11}{36}\approx0.306$.

```{r}
mean(eventE)  # P(E)
mean(eventF)  # P(F)
```

To estimate $P(E \cap F) = \frac{2}{36} \approx 0.056$:

```{r}
mean(eventE & eventF)
```

To estimate $P(E \cup F) \approx 0.389$:

```{r}
mean((sumDice == 6) | (die1 == 2 | die2 == 2))
```

**Bernouilli's theorem:**

Let $E$ be an event with probability $p$. Let $m_n$ be the number of times that $E$ occurs in $n$ repeated trials, where we assume the outcome of trials do not affect the outcome of other trials. Then

$$
\lim_{x\to\infty} \frac{m_n}{n}=p
$$

## 2.2.3 Using `replicate` to repeat experiments
