---
title: "Chapter 3 Discrete Random Variables"
format: html
editor: visual
---

Let $S$ be the sample space of an experiment. A *random variable* is a function from $S$ to the real line. Random variables are usually denoted by a capital letter. Many times we will abbreviate the words *random variable* with *rv*.

Example: three coins are tossed. The sample space is

$$
S=\{HHH,HHT,HTH,HTT,THH,THT,TTH,TTT\}
$$

All eight outcomes are equally likely, each occuring with a probability 1/8.

A natural random variable here is the number of heads observed, which we call $X$. As a function from $S$ to real numbers, $X$ is given by:

$$
X(HHH)=3
$$

$$
X(HHT)=X(HTH)=X(THH)=2
$$

$$
X(TTH)=X(THT)=X(HTT)=1
$$

$$
X(TTT)=0
$$

The event $X=2$ is the set of outcomes $\{HHT,HTH,THH\}$ and so:

$$
P(X=2)=P(\{HHT,HTH,THH\})=\frac{3}{8}
$$

# 3.1 Probability mass functions

A *discrete* random variable is a random variable that takes integer values. A discrete random variable is characterized by its *probability mass function* (pmf). The pmf $p$ of a random variable $X$ is given by

$$
p(x)=P(X=x)
$$

Let $p$ be the pmf of $X$:

1.  $p(x) \geq 0$ for all $x$
2.  $\sum_x p(x) = 1$

Example 3.4

```{r}
litterpmf <- c(0.18, 0.51, 0.27, 0.04)
sample(x = 1:4, size = 30, replace = TRUE, prob = litterpmf)
```

```{r}
X <- sample(1:4, 10000, replace = TRUE, prob = litterpmf)
mean(X > 1)
```

```{r}
table(X) / 10000
```

Example 3.5

```{r}
coin_toss <- sample(c("H", "T"), 3, replace = TRUE)
sum(coin_toss == "H")
```

```{r}
X <- replicate(10000, {
  coin_toss <- sample(c("H", "T"), 3, replace = TRUE)
  sum(coin_toss == "H")
})
head(X, 30)
table(X) / 10000
```

Example 3.6

```{r}
X <- replicate(10000, {
  coin_toss <- sample(c("H", "T"), 3, replace = TRUE)
  sum(coin_toss == "H")
})
mean(X >= 1)
```

# 3.2 Expected value

For a discrete random variable $X$ with *pmf* $p$, the *expected value* of $X$ is

$$
E[X]=\sum_x xp(x)
$$

provided this sum exists, where the sum is taken over all possible values of the random variable $X$.

Law of large numbers: the mean of $n$ observations of a random variable $X$ converges to the expected value $E[X]$ as $n \rightarrow \infty$, assuming $E[X]$ is defined.

```{r}
# Example 3.9
rolls <- sample(1:6, 30, replace = TRUE)
rolls
mean(rolls)

rolls <- sample(1:6, 10000, replace = TRUE)
mean(rolls)
```

The mean value is balanced halfway between 1 and 6, at 3.5

```{r}
# Example 3.11
# Let X be the number of kittens in a Eurasian lynx litter.
litterpmf <- c(0.18, 0.51, 0.27, 0.04)
sum((1:4) * litterpmf)  # 2.17 kittens on average

X <- sample(1:4, 10000, replace = TRUE, prob = litterpmf)
mean(X)
```

# 3.3 Binomial and geometric random variables

A *Bernouilli trial* is an experiment that can result in two outcomes, which we will denote as "success" and "failure". The probability of a success will be denoted $p$, and the probability of failure is therefore $1-p$.

## 3.3.1 Binomial

A random variable $X$ is said to be a *binomial random variable* with parameters $n$ and $p$ if

$$
P(X=x)={n \choose x}p^x(1-p)^{n-x}
$$

```{r}
# Number of heads when three coins are tossed
x <- 0:3
dbinom(x, size = 3, prob = 0.5)
```

Let $X$ be a *binomial random variable* with $n$ trials and probability of success $p$. Then

$$
E[X]=np
$$

```{r}
# 100 dice are thrown. 
# What is the expected number of sixes?
100 * (1/6)

# What is the probability of observing 10 or fewer sixes?
# 0 sixes:
dbinom(x = 0, size = 100, prob = 1/6)
# 1 six:
dbinom(x = 1, size = 100, prob = 1/6)
# 10 or less sixes:
sum(dbinom(x = 0:10, size = 100, prob = 1/6))
```

`pbinom` is the cumulative sum of the pmf:

```{r}
pbinom(q = 10, size = 100, prob = 1/6)
```

```{r}
# Peyton and Riley are running for office, and 46% of all voters
# prefer Peyton. A poll randomly selects 300 voters.

# What is the expected number of voters who will report a preference for
# Peyton?
0.46 * 300

# What is the probability that the poll results suggest Peyton will win?
1 - pbinom(q = 150, size = 300, prob = 0.46)
```

```{r}
# 15 simulations of the Peyton vs. Riley poll
rbinom(n = 15, size = 300, prob = 0.46)
```

```{r}
# Compute P(X > 150) by simulation
X <- rbinom(n = 10000, size = 300, prob = 0.46)
mean(X > 150)
```

## 3.3.2 Geometric

A random variable $X$ is said to be a *geometric random variable* with parameter $p$ if

$$
P(X=x)=(1-p)^x p
$$

Let $X$ be the random variable that counts the number of failures before the first success in a Bernouilli process with probability of success $p$. Then $X$ is a geometric random variable.

```{r}
# A die is tossed until the first 6 occurs. What is the probability that it takes 4 or more tosses?
sum(dgeom(x = 3:1000, prob = 1/6))
1 - sum(dgeom(x = 0:2, prob = 1/6))
1 - pgeom(q = 2, prob = 1/6)
pgeom(q = 2, prob = 1/6, lower.tail = FALSE)
```

```{r}
# Simulation
X <- rgeom(n = 10000, prob = 1/6)
mean(X >= 3)
```

Let $X$ be a geometric random variable with probability of success $p$. Then

$$
E[X]=\frac{(1-p)}{p}
$$

# 3.4 Functions of a random variable
