---
title: "Chapter 3 Discrete Random Variables"
format: html
editor: visual
---

Let $S$ be the sample space of an experiment. A *random variable* is a function from $S$ to the real line. Random variables are usually denoted by a capital letter. Many times we will abbreviate the words *random variable* with *rv*.

Example: three coins are tossed. The sample space is

$$
S=\{HHH,HHT,HTH,HTT,THH,THT,TTH,TTT\}
$$

All eight outcomes are equally likely, each occuring with a probability 1/8.

A natural random variable here is the number of heads observed, which we call $X$. As a function from $S$ to real numbers, $X$ is given by:

$$
X(HHH)=3
$$

$$
X(HHT)=X(HTH)=X(THH)=2
$$

$$
X(TTH)=X(THT)=X(HTT)=1
$$

$$
X(TTT)=0
$$

The event $X=2$ is the set of outcomes $\{HHT,HTH,THH\}$ and so:

$$
P(X=2)=P(\{HHT,HTH,THH\})=\frac{3}{8}
$$

# 3.1 Probability mass functions

A *discrete* random variable is a random variable that takes integer values. A discrete random variable is characterized by its *probability mass function* (pmf). The pmf $p$ of a random variable $X$ is given by

$$
p(x)=P(X=x)
$$

Let $p$ be the pmf of $X$:

1.  $p(x) \geq 0$ for all $x$
2.  $\sum_x p(x) = 1$

Example 3.4

```{r}
litterpmf <- c(0.18, 0.51, 0.27, 0.04)
sample(x = 1:4, size = 30, replace = TRUE, prob = litterpmf)
```

```{r}
X <- sample(1:4, 10000, replace = TRUE, prob = litterpmf)
mean(X > 1)
```

```{r}
table(X) / 10000
```

Example 3.5

```{r}
coin_toss <- sample(c("H", "T"), 3, replace = TRUE)
sum(coin_toss == "H")
```

```{r}
X <- replicate(10000, {
  coin_toss <- sample(c("H", "T"), 3, replace = TRUE)
  sum(coin_toss == "H")
})
head(X, 30)
table(X) / 10000
```

Example 3.6

```{r}
X <- replicate(10000, {
  coin_toss <- sample(c("H", "T"), 3, replace = TRUE)
  sum(coin_toss == "H")
})
mean(X >= 1)
```

# 3.2 Expected value
